---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Activity Scope

- This activity depends on the existence of a dataset called **Activity Monitoring Data**. if you don't have one, you could download it from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). meanwhile, this **R Markdown** handles this issues for you, ***Just Enjoy :)***.

```{r fileValidation, echo=FALSE}

# URL for the required data sets file

dataSetsURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

# Checking if the dataset exists, if not download it

if(!file.exists("activity.csv")){
    
    if(!file.exists("activity.zip")){
        
        # Downloading the data sets file
        
        download.file(url = dataSetsURL, destfile = "activity.zip")
        
    }
    
}

# Unzipping the downloaded data sets file

unzip(zipfile = "activity.zip", overwrite = TRUE)

```

## Loading and preprocessing the data

```{r dataPrep}

# Importing the required dataset

dataset <- read.csv(file = "activity.csv")

# Dataset sample preview

head(dataset)

```

## What is mean total number of steps taken per day?

**1. aggregating the total number of steps per day**

```{r totalPerDay}

# Calculating the total number of steps taken per day

total_steps_pd <- aggregate(steps ~ date, subset(dataset, steps != "NA"), sum)

# Dataset sample preview

head(total_steps_pd)

```

**2. Histogram of the total number of steps taken each day**

```{r histTotalPerDay, fig.width=10, fig.height= 6, warning=FALSE, error=FALSE}

library(ggplot2)

ggplot(total_steps_pd, aes(total_steps_pd$steps)) +  
geom_histogram(breaks=seq(0, 25000, by=5000), col = "dodgerblue4",fill="dodgerblue4") +  
theme(axis.title.x = element_text(colour = "dodgerblue4", face = "bold"),
      axis.text.x = element_text(colour = "dodgerblue4"),
      axis.title.y = element_text(colour = "firebrick2", face = "bold"),
      axis.text.y = element_text(colour = "firebrick2"),
      legend.text = element_text(colour = "firebrick2"),
      legend.title = element_text(colour = "dodgerblue4", face = "bold"),
      plot.title = element_text(hjust = 0.5, colour = "dodgerblue4", face = "bold")) + 
scale_y_continuous(breaks = seq(0, 30, by=5)) +  
xlab("Steps") +  
ylab("Frequency") + 
labs(fill = "Steps") +  
ggtitle("Histogram of The Total Number of Steps Taken Each Day")

```

**3. Answring the question of What is mean total number of steps taken per day?**  

```{r bMeanMedTotal, results='hide'}

bmean <- as.integer(round(mean(total_steps_pd$steps),0))
bmedian <- as.integer(round(median(total_steps_pd$steps),0))

```

+  **So, the Mean is : `r bmean`, and the Median is : `r bmedian`**

## What is the average daily activity pattern?

**1. Calculating the average number of steps taken across all days grouped by *the 5-minute interval***

```{r avgPerInterval}

# Calculating the average number of steps taken across all days by interval

avg_steps_pi <- aggregate(steps ~ interval, subset(dataset, steps != "NA"), mean)

# Dataset sample preview

head(avg_steps_pi)

```

**2. Time series plot of the average number for all intervals**

```{r tsAvgPerInterval, fig.width=8, fig.height= 5, warning=FALSE, error=FALSE}

ggplot(avg_steps_pi, aes(interval, steps, fill = steps)) +  
geom_line(color="royalblue4", size = 1, aes(fill = steps), position = "dodge", stat="identity") +  
theme(axis.title.x = element_text(colour = "dodgerblue4", face = "bold"),
      axis.text.x = element_text(colour = "dodgerblue4"), 
      axis.title.y = element_text(colour = "firebrick2", face = "bold"),
      axis.text.y = element_text(colour = "firebrick2"),
      plot.title = element_text(hjust = 0.5, colour = "dodgerblue4", face = "bold")) +  
xlab("Intervals") +  
ylab("Average Steps Across All Days") +  
ggtitle("Time Series Plot of The Average Number of All Intervals Across All Days")

```

**3. Answring the question of Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**  

```{r maxInterval, results='hide'}

maxInterval = subset(avg_steps_pi, steps == max(avg_steps_pi$steps))[,1]

```

+ **So, the 5-minute interval contains the maximum number of steps is : `r maxInterval`.** 

## Imputing missing values

**1. Calculating the total number of missing values in the Activity Monitoring Data :**  

```{r nas, results='hide'}

nas_ds <- sum(is.na(dataset$steps))

```

+ **So, there is `r nas_ds` missing values in the Activity Monitoring Data.** 

**2. Devising a strategy for filling in all of the missing values in the Activity Monitoring Data :**  

```{r strategyCHK}

# Creating new dataset to decide which strategy to apply, other words, filling NAs with the mean/median for that day, or the mean for that 5-minute interval

chk_dataset <- data.frame(date = dataset$date, 
                          interval = dataset$interval, 
                          is_na = is.na(dataset$steps),
                          not_na = !is.na(dataset$steps))

# Dataset sample preview

head(chk_dataset)

```

+ **a) Summarizing By Date :**  

```{r strategyBD, warning=FALSE, error=FALSE}

library(plyr)

# Creating a new aggregated dataset grouped by Date

chk_by_date <- ddply(chk_dataset, .(date), summarise, is_na = sum(is_na), not_na = sum(not_na))

chk_by_date_no <- nrow(chk_by_date)

# Dataset sample preview

head(chk_by_date)

```

+ **b) Challenging Date :**  

```{r strategyBDChal, results='hide', warning=FALSE, error=FALSE}

# Identifying the weight of NAs values across all days

strategyBD <- nrow(subset(chk_by_date, is_na != 0 & not_na != 0))

```

+ **So, There is `r strategyBD` rows out of `r chk_by_date_no` total rows in the aggregated dataset grouped by Date that we could replace NAs with.**

+ **c) Summarizing By Interval :**  

```{r strategyBI, warning=FALSE, error=FALSE}

# Creating a new aggregated dataset grouped by Interval

chk_by_interval <- ddply(chk_dataset, .(interval), summarise, is_na = sum(is_na), not_na = sum(not_na))

chk_by_interval_no <- nrow(chk_by_interval)

# Dataset sample preview

head(chk_by_interval)

```

+ **d) Challenging Interval :**  

```{r strategyBIChal, results='hide', warning=FALSE, error=FALSE}

# Identifying the weight of NAs values across all days

strategyBI <- nrow(subset(chk_by_interval, is_na != 0 & not_na != 0))

```

+ **So, There is `r strategyBI` rows out of `r chk_by_interval_no` total rows in the aggregated dataset grouped by Interval that we could replace NAs with.**

**Obviously from the above, we tend to fill in all of the missing values in the Activity Monitoring Data using the mean for each 5-minute interval across all the days.**  

**3. Creating a new dataset that is equal to the original dataset but with the missing data filled in :**  

```{r paralData, warning=FALSE, error=FALSE}

# Merging the original dataset with the dataset containing the average number for all intervals

filled_dataset <- merge(dataset, avg_steps_pi, by = "interval", suffixes = c("_org","_avg"))

# Creating new column holds the original steps value and used to replace NAs in it to help tracing the changes

filled_dataset$steps <- filled_dataset$steps_org

# Filling in the NAs with its intervals' mean 

filled_dataset[is.na(filled_dataset$steps), "steps"] <- filled_dataset[is.na(filled_dataset$steps), "steps_avg"]

# Arranging and reordering the new dataset

filled_dataset <- filled_dataset[order(filled_dataset$date,filled_dataset$interval), c("steps","date","interval","steps_org","steps_avg")]

row.names(filled_dataset) <- c(1:nrow(filled_dataset))

# Dataset sample preview

head(subset(filled_dataset, is.na(steps_org)), 10)

```

**4. Making a histogram of the total number of steps taken each day after filling the NAs across al days :** 

+ **a) Aggregating the total number of steps per day**

```{r totalFillPerDay}

# Calculating the total number of steps taken per day

total_fill_steps_pd <- aggregate(steps ~ date, subset(filled_dataset, steps != "NA"), sum)

# Dataset sample preview

head(total_fill_steps_pd)

```

+ **b) Creating a new dataset contains all the total steps per day before and after filling NAs :**

```{r totalPerDayBA}

# Adding new column to each dataset to identify the source on the plot

total_steps_pd$type <- c("Before Filling NAs")

total_fill_steps_pd$type <- c("After Filling NAs")

# Calculating the total number of steps taken per day

total_steps_pd_ba <- rbind(total_steps_pd, total_fill_steps_pd)

# Dataset sample preview

head(total_steps_pd_ba)

```

+ **c) Histogram of the total number of steps taken each day before and after filling in NAs :**

```{r histTotalPerDayBA, fig.width=10, fig.height= 6, warning=FALSE, error=FALSE}

ggplot(total_steps_pd_ba, aes(total_steps_pd_ba$steps, fill = total_steps_pd_ba$type)) +  
geom_histogram(breaks=seq(0, 25000, by=5000),position = "identity") +  
theme(axis.title.x = element_text(colour = "dodgerblue4", face = "bold"),
      axis.text.x = element_text(colour = "dodgerblue4"),
      axis.title.y = element_text(colour = "firebrick2", face = "bold"),
      axis.text.y = element_text(colour = "firebrick2"),
      legend.text = element_text(colour = "firebrick2"),
      legend.title = element_text(colour = "dodgerblue4", face = "bold"),
      plot.title = element_text(hjust = 0.5, colour = "dodgerblue4", face = "bold")) + 
scale_y_continuous(breaks = seq(0, 35, by=5)) +  
scale_fill_manual(values = c("firebrick2","dodgerblue4")) +  
xlab("Steps") +  
ylab("Frequency") + 
labs(fill = "Steps") +  
ggtitle("Histogram of The Total Number of Steps Taken Each Day Before/After Filling NAs")

```

+ **d) Answring the question of Do these values differ from the estimates from the first part of the assignment?**  

```{r aMeanMedTotal, results='hide'}

amean <- as.integer(round(mean(total_fill_steps_pd$steps),0))
amedian <- as.integer(round(median(total_fill_steps_pd$steps),0))
asum <- as.integer(round(sum(total_fill_steps_pd$steps),0))

```

**So, the mean before filling NAs was : `r bmean`, while after filling NAs become : `r amean`,**  

**While, the median before filling NAs was : `r bmedian`, and after filling NAs become : `r amedian`**  

+ **Almostly, There is no difference in the mean and median total number of steps taken per day before and after filling in the NAs.**  

+ **e) Answring the question of What is the impact of imputing missing data on the estimates of the total daily number of steps?**  

```{r baTotal, results='hide'}

bsum <- as.integer(round(sum(total_steps_pd$steps),0))
asum <- as.integer(round(sum(total_fill_steps_pd$steps),0))
vsum <- as.integer(round(100 - ((bsum / asum) * 100),0))

```

**So, the total daily number of steps before filling NAs was : `r bsum`, while after filling NAs become : `r asum`,**  

+ **Obviously, There is increase in the total number of steps taken per day before and after filling in the NAs by : `r vsum`%.** 

## Are there differences in activity patterns between weekdays and weekends?

**1. Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day :**  

```{r dayType}

# Creating vector holds the weekend days

weekends <- c('Saturday', 'Sunday')

# Creating a factor to identify the day type and assign it to a new column in the filled dataset

filled_dataset$day_type <- factor((weekdays(as.Date(filled_dataset$date)) %in% weekends), 
                                  levels = c(TRUE, FALSE), 
                                  labels = c("weekend","weekday"))

# Dataset sample preview

head(filled_dataset, 10)

```

**2. Creating a new dataset containing the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days :** 

```{r avgIntByDT, warning=FALSE, error=FALSE}

# Creating a new aggregated dataset grouped by interval across all weekdays and weekends

avgIntByDT <- ddply(filled_dataset, .(interval, day_type), summarise, steps = mean(steps))

# Dataset sample preview

head(avgIntByDT, 10)

```

**3. Making a panel plot containing a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all day types (y-axis)**  

```{r tsAvgIntByDT, fig.width=8, fig.height= 8, warning=FALSE, error=FALSE}

ggplot(avgIntByDT, aes(interval, steps, fill = steps)) +  
geom_line(color="royalblue4", size = 1, aes(fill = steps), position = "dodge", stat="identity") +  
theme(axis.title.x = element_text(colour = "dodgerblue4", face = "bold"),
      axis.text.x = element_text(colour = "dodgerblue4"), 
      axis.title.y = element_text(colour = "firebrick2", face = "bold"),
      axis.text.y = element_text(colour = "firebrick2"),
      plot.title = element_text(hjust = 0.5, colour = "dodgerblue4", face = "bold")) +  
facet_wrap( ~ day_type, nrow = 2) +  
xlab("Intervals") +  
ylab("Average Steps Across All Day Types") +  
ggtitle("Time Series Plot of The Average Number of All Intervals Across All Day Types")

```
  
  
  
  
  
